\chapter{File Formats}
\section{AKY}
Music file
\section{BAS}
Basic file
\section{BMP}
Windows bitmap. On the ZX Spectrum next, these must be uncompressed
and 8 bpp.
\section{DSK}
Disk image
\section{ESP}
Update file for Wi-Fi module
\section{GDE}
ZX Spectrum Next Guide hypertext file
\section{MC}
A Timex hi-color image. Like .shc, but the bitmap and colour map are
stored in display order.
\section{MLT}
A Timex hi-color image. Like .shc, but the colour map is stored in
display order.
\section{MOD}
Modtracker music file (requires Pi accelerator to play).
\section{NDR}
Nexdaw music file
\section{NEX}
ZX Spectrum Next executable file.
\section{NXI}
A version of a 256x192x8 layer 2 image.
\begin{itemize}
\item bytes 0-511 - palette data
\item bytes 512-49663 - image data
\end{itemize}
\section{O}
ZX-80 executable file
\section{P3D}
Disk image file
\section{PT3}
Protracker 3 music file
\section{P}
ZX-81 executable file
\section{SDH}
Music file (requires Pi accelerator to play)
\section{SID}
Commodore 64 SID music file (requires Pi accelerator to play)
\section{SCR}
ZX Spectrum screen dump

In basic form a 6912 byte file containing the bitmap and colour map
for a standard spectrum image. May have ULAPlus or ULANext color data
appended.

ULAplus
\begin{itemize}
\item bytes 0-6143 - bitmap: memory order
\item bytes 6144-6911 - colour map: memory order
\item bytes 6912-6975 - ULAPlus palette data
\end{itemize}
ULANext
\begin{itemize}
\item bytes 0-6143 - bitmap: memory order
\item bytes 6144-6911 - colour map: memory order
\item byte 6912 - ULANext mode
\item bytes 6913-between 6228 and 7426 - ULANext palette data, size is
  dependent on specific ULANext mode and 8 or 9 bit palette entries.
  \begin{tabular}{l | r | r | r | r}
    mode & ink & paper & 8-bit & 9-bit \\
    \$01 & 2 & 128 & 130 & 260\\
    \$03 & 4 & 64 & 68 & 136\\
    \$07 & 8 & 32 & 40 & 80\\
    \$0F & 16 & 16 & 32 & 64\\
    \$1F & 32 & 8 & 40 & 80\\
    \$3F & 64 & 4 & 68 & 136\\
    \$7F & 128 & 2 & 130 & 260\\
    \$FF & 256 & 1 & 257 & 514
  \end{tabular}
\end{itemize}
\section{SHC}
Timex hi-color screen dump

In basic form the 12,288 byte file containing the bitmap and colour
map for a Timex hi-color image. May have ULAPlus or ULANext color data
appended.

ULAplus
\begin{itemize}
\item bytes 0-6143 - bitmap: memory order
\item bytes 6144-12287 - colour map: memory order
\item bytes 12288-12351 - ULAPlus palette data
\end{itemize}
ULANext
\begin{itemize}
\item bytes 0-6143 - bitmap: memory order
\item bytes 6144-12287 - colour map: memory order
\item byte 12288 - ULANext mode
\item bytes 12289-between 12304 and 12802 - ULANext palette data, size
  is dependent on specific ULANext mode and 8 or 9 bit palette entries.
  \begin{tabular}{l | r | r | r | r}
    mode & ink & paper & 8-bit & 9-bit \\
    \$01 & 2 & 128 & 130 & 260\\
    \$03 & 4 & 64 & 68 & 136\\
    \$07 & 8 & 32 & 40 & 80\\
    \$0F & 16 & 16 & 32 & 64\\
    \$1F & 32 & 8 & 40 & 80\\
    \$3F & 64 & 4 & 68 & 136\\
    \$7F & 128 & 2 & 130 & 260\\
    \$FF & 256 & 1 & 257 & 514
  \end{tabular}
\end{itemize}
\section{SHR}
  
Timex hi-resolution screen dump.

A 12,288 or 12,289 byte file containing a memory dump of the two 6k
screen areas used by Timex hi-resolution mode in memory order with an
optional byte indicating teh colours used.
\section{SL2}
ZX Spectrum Next layer 2 screen dump

In its most basic standard form, this is a 49,152 byte file contaning
the memory for the image data. 

256x192x8
\begin{itemize}
\item bytes 0-49151 - image data: row dominant
\item optional bytes 49152-49407 or 49663 - palette data (256 bytes
  for 8-bit entries, 512 bytes for 9-bit entries)
\end{itemize}
320x256x8
\begin{itemize}
\item bytes 0-81919 - image data: column dominant
\item optional bytes 81920-82175 or 82431 - palette data (256 bytes
  for 8-bit entries, 512 bytes for 9-bit entries)
\end{itemize}
640x256x4
\begin{itemize}
\item bytes 0-81919 - image data: column dominant, nybbles in adjacent columns
\item optional bytes 81920-81935 or 81951 - palette data (16 bytes for
  8-bit entries, 32 bytes for 9-bit entries)
\end{itemize}
\section{SLR}
ZX Spectrum Next low-resolution screen dump.

Basic standard form 12,288 byte file contaning the memory for the two
6k areas for 128x96x8 image files

192x96x8
\begin{itemize}
\item bytes 0-12287 - image data: row dominant
\item optional bytes 12288-12543 or 12799 - palette data (256 bytes
  for 8-bit entries, 512 bytes for 9-bit entries)
\end{itemize}
192x96x4 (Radistan)
\begin{itemize}
\item bytes 0-6143 - image data: row dominant
\item optional bytes 6144-6159 or 6175 - palette data (16 bytes for
  8-bit entries, 32 bytes for 9-bit entries)
\end{itemize}
\section{SNA}
ZX Spectrum memory snapshot
\section{SND}
Music file (requires Pi accelerator to play)
\section{SNX}
Extended ZX Spectrum memory snapshot to be run in ZX Spectrum Next mode
\section{SPR}
Sprite file
\section{TAP}
ZX Spectrum tape image
\section{TXT,DOC,ASM,INI,CFG,MD}
Text files
\section{TZX}
ZX Spectrum tape image (able to handle custom tape formats). Requires
Pi accelerator to load.
\section{WAV}
PCM audio file
\section{XM}
XM tracker music file (requires Pi accelerator to play)
\section{Z3 to Z8}
Inform interactive fiction files
\section{Z80}
ZX Spectrum memory snapshot
\section{ZIP}
Zip compressed file
