Tile-sprite configuration (TSconf). Introduction

TS-Conf is a modern extension of the ZX Spectrum, which brings
long-awaited and features in the form of pixel oriented color,
hardware sprites, etc.

Graphic subsystem

Tao says: The screen we see is the output window of the resolution
specified in the system, which displays a block of 512x512 video
memory pages in the specified mode according to the given output
coordinates. TSU is a device that collects data from video, tile and
sprite memory, which processes and presents it at the current position
of the line drawing in a given resolution / desired mode.

This unit is an extension (add-on) above the standard 6912 byte
Spectrum screen. More precisely, the 6912 byte mode is part of the
whole family of system resolutions:
\begin{itemize}
\item 256x192
\item 320x200
\item 320x240
\item 360x288
\end{itemize}

These permissions can be used in different video modes:
\begin{itemize}
\item ZX
\item 16c
\item 256c
\item Text
\end{itemize}

To display the color for a given mode, the first block of the system's
internal memory is used - the palette memory . This block represents
512 bytes and stores the full color palette of the system - two-byte
data of color components for 256 colors.

The full palette is divided into groups of 16 colors (32 bytes) for
16c mode, which gives us 16 palettes with 16 colors.

The first palette is number 0. 

When you turn on the system, a set of 16 colors for the ZX mode is
loaded into the general palette, which is located in the last, 15th
palette.

Each mode sets its own characteristics of displaying color information
on the screen:
\begin{itemize}
\item ZX limits the output to the 16 colors of the standard ZX
  Spectrum palette;
\item 16c - in this mode only 16 colors are used - one of the 16
  available palettes;
\item 256c - provides the ability to display all the loaded colors in
  the palette
\item Text - text mode, allows you to display text in color.
\end{itemize}

In this case, it was a question of horizontal mobility. Let's talk
about vertical:

In addition to the listed modes, the system allows the use of layered
display of graphical information. The standard Spectrum screen is the
base screen, but it is not the lowest.

So, the location of the layers that form the screen: 
\begin{itemize}
\item[] 1. Border. A monochromatic fullscreen layer. Set the color of
  the border
\item[] 2. Basic (main) screen. May be included in any of the
  resolution/mode combinations
\item[] 3. Accelerated graphics layers displayed in the resolution
  specified for the base screen in 16 color mode
  \begin{itemize}
  \item sprite layer 0 
  \item tile layer 0 
  \item sprite layer 1 
  \item tile layer 1 
  \item sprite layer 2
  \end{itemize}
\end{itemize}

Thus, we have 7 layers that make up a single, visible screen.

All the layers use the colors specified in a common palette of 256
colors.

The tile and sprite layers only work in 16c mode, the 0th color of
each palette set is transparent. Thanks to TSU, there is no need to
save video memory data under tiles and sprites, since the display on
the screen is literally collected at the output of each line without
changing the contents of the video memory.

Each graphic element of these two types of layers represents a block
of at least 8x8 pixels, and for each of them it is necessary to
specify a palette, any one the available 16.

The tile layer is a map describing the location of the graphic
elements specified as an image. Thus, the position of the tile on the
screen directly depends on its position in the map.

Tile patterns: 64x64 tiles (4096 tiles in total), uses up to 4
palettes from 4 groups of all 16 palettes can be used for one
layer. For each tile it is possible to set its own (out of these 4)
palette.

In total, we have two such layers organized in the same way. 

Sprites are graphics, organized like tiles, but having large (multiple
of 8) sizes - from 8x8 to 64x64 pixels. Sprites also have zero color
transparency.

A feature of the sprites is that for each sprite, you can specify both
your its own palette and position to the pixel resolution.

The system uses the second block of internal memory to handle sprites,
representing the next 512 bytes of memory, which stores "sprite
handles"- data from 6 bytes, describing each sprite. The maximum
number of descriptors in this memory is 85 pieces.

In the following articles I will describe in more detail about working
with the devices listed at the beginning of the article.

PS: for those who do not want to wait and want details
