\section{Layer 2}
Layer 2 is a for bitmapped graphics. It supports modes with
$256\times192\times256$ resolution at global coordinates (32, 32) to
(287, 223) mapped linearly left to right/top to bottom,
$320\times256\times256$ resolution at global coordinates (0, 0) to
(318, 255) mapped top to bottom/left to right, and
$640\times256\times16$ resolution at global coordinates (0, 0) to
(319, 255) with half width pixels mapped so that the nybbles in a byte
are adjacent columns (MSN on the left) and bytes running top to
bottom/left to right. It can be mapped starting at any 16k memory
blocks. The $256\times192\times256$ mode requires 3 consecutive blocks
(48k) while the others use 5 consecutive blocks (80k).

\subsection{Configuration}
Layer 2 is enabled using port \$123B or register \$69. The mode is
selected using register \$70. How layer 2 memory is overlaid on main
memory is controled by port \$123B and register \$70. The location in
memory is controlled by register \$12 with a shadow area pointed to by
register \$13 for double buffering. Finally port \$123B is used to
select either the main RAM area or the shadow RAM area for rendering
the layer.

\input{ports/123B.tex}
\input{registers/tbblue/12.tex}
\input{registers/tbblue/13.tex}
\input{registers/tbblue/69.tex}
\input{registers/tbblue/70.tex}

\paragraph{Layer 2 $256\times192$, Write only overlaid on ROM}
\begin{verbatim}
p_layer2: defl $123b
start:
  ld bc,p_layer2
  ld a,$03       ; enable, wo, 1st 16k
  out (c),a
  call wrtpage
  ld bc,p_layer2
  ld a,$43       ; enable, wo, 2nd 16k
  out (c),a
  call wrtpage
  ld bc,p_layer2
  ld a,$83       ; enable, wo, 3rd 16k
  out (c),a
  call wrtpage
  ret
wrtpage:  
  ld hl,$0000
  ld bc,$0040    ; 40*256 writes
loop:
  ld (hl),b
  inc hl
  djnz loop
  dec c
  jr nz,loop
\end{verbatim}

\paragraph{Layer 2 $256\times192$ resolution}
\begin{verbatim}
r_mmu_7:  defl $57
r_disp1:  defl $69
r_layer2: defl $70
start:
  nextreg r_disp1,$80  ; enable layer 2
  nextreg r_layer2,$00 ; 256x192x256
  ld a,$12             ; page 18=bank 9
loop1:
  nextreg r_mmu_7,a    ; map page into slot 7
  ld bc,$0020          ; 20*256 = 8k
  ld hl,$E000          ; address of slot 7
loop2:
  ld (hl),b
  inc hl
  djnz loop2
  dec c
  jp NZ,loop2
  inc a
  cp $18               ; stop at page 24
  jp NZ,loop1
\end{verbatim}

\paragraph{Layer 2 $320\times256$ resolution}
\begin{verbatim}
r_mmu_7:  defl $57
r_disp1:  defl $69
r_layer2: defl $70
start:
  nextreg r_disp1,$80  ; enable layer 2
  nextreg r_layer2,$10 ; 320x256x256
  ld a,$12             ; page 18=bank 9
loop1:
  nextreg r_mmu_7,a    ; map page into slot 7
  ld bc,$0020          ; 20*256 = 8k
  ld hl,$E000          ; start of slot 7
loop2:
  ld (hl),b
  inc hl
  djnz loop2
  dec c
  jp NZ,loop2
  inc a
  cp $1C               ; stop at page 28
  jp NZ,loop1
\end{verbatim}

\paragraph{Layer 2 $640\times256$ resolution}
\begin{verbatim}
r_mmu_7:  defl $57
r_disp1:  defl $69
r_layer2:  defl $70
start:
  nextreg r_disp1, $80   ; enable layer 2
  nextreg r_layer2, $20  ; 640x256x16
  ld a, $12    ; page 18=bank 9
loop1:
  nextreg r_mmu_7, a  ; map page into slot 7
  ld bc, $0020    ; 20*256 = 8k
  ld hl, $E000    ; start address for slot 7
loop2:
  ld (hl), b
  inc hl
  djnz loop2
  dec c
  jp NZ, loop2
  inc a
  cp $1C      ; stop at page 28
  jp NZ, loop1
\end{verbatim}

\subsection{Scrolling}
Scrolling Layer 2 is controlled by registers \$16 and \$17. (Is there
a third scrolling register for layer 2?)

\input{registers/tbblue/16.tex}
\input{registers/tbblue/17.tex}

\subsection{Clipping}
The Clip area for is based on the local coordinate system for the mode
in question and is set using register \$18 with the option of
selection which write in active using register \$1C.

\input{registers/tbblue/18.tex}
\input{registers/tbblue/1C.tex}
