\chapter{Interrupts}

\section{Interrupt Types}
The Z80 has three different hardware interrupt signals:
\textoverline{RESET}, \textoverline{NMI}, and \textoverline{INT}.

\paragraph{\textoverline{RESET}}
\textoverline{RESET} is used to return the CPU to a known state. When
the \textoverline{RESET} line is pulled low, a \textoverline{RESET} is
generated. The CPU then does several things. I, and R are set to \$00.
PC is set to \$0000. SP becomes \$FFFF. A and F are set to \$FF.  The
interrupt mode is set to 0. And (maskable) interrupts are disabled by
clearing IFF1 and IFF2.

\paragraph{\textoverline{NMI}}
\textoverline{NMI} is the non-maskable interrupt. Upon receiving a
non-maskable interrupt (\textoverline{NMI} being pulled low) one of
two sequences occur depending on the calue of bit 3 of the interrupt
control register (nextreg \$C0).

\input{registers/tbblue/C0.tex}

If bit 3 is clear (0) PC is pushed on the stack, IFF1 is copied to
IFF2, IFF1 is cleared (inhibiting maskable interrupts). The
\textoverline{NMI} should end with RETN which copies the contents of
IFF2 to IFF1 (returning the interrupt state to what it was before the
\textoverline{NMI}) and PC is popped off the stack.

If bit 3 is set (1) PC is stored in the \textoverline{NMI} return
address registers (nextregs \$C2 and \$C3), IFF1 is copied to IFF2,
IFF1 is cleared (inhibiting maskable interrupts). The
\textoverline{NMI} should end with RETN which copies the contents of
IFF2 to IFF1 (returning the interrupt state to what it was before the
\textoverline{NMI}) and PC is compied from the \textoverline{NMI}
return address registers.

\input{registers/tbblue/C2.tex}
\input{registers/tbblue/C3.tex}

\paragraph{\textoverline{INT}}
The interrupt generally of most interest to programmers is
\textoverline{INT}. So much so that if programmers talk about
interrupts on the Z80, they are probebly only talking about
\textoverline{INT}. The processing of \textoverline{INT} is controlled
by IFF1 and IFF2 which are set using EI to enable interrupts and reset
using DI to disable interrupts. Interrupts can happen at any time and
should preserve register contents.  If none of your code uses the
alternate registers the EXX and EX AF,AF’ instructions can make this
faster and easier. Interrupt routined should end with EI and RETI to
reenable interrupts, potentially inform the interrupting device that
its interrupt has been serviced, and return from the interrupt
routine. In general the Spectrum machines do not make any distingtion
between RET and RETI, but future developments in the ZX Spectrum Next
may make the distinction important.

The ZX Spectrum Next has 14 internal sources for \textoverline{INT}
signals. This can be enabled and disabled using nextregs \$C4 --
\$C6. Which signals have been received can be read/cleared using
nexregs \$C8 -- \$CA.

Interrupt Enable

\input{registers/tbblue/C4.tex}
\input{registers/tbblue/C5.tex}
\input{registers/tbblue/C6.tex}

Interupt Status

\input{registers/tbblue/C8.tex}
\input{registers/tbblue/C9.tex}
\input{registers/tbblue/CA.tex}

Internal Interrupt Sources
\begin{itemize}
\item[] 0 = Line (highest priority)
\item[] 1 = UART 0 Rx
\item[] 2 = UART 1 Rx
\item[] 3-10 = CTC channels 0-7
\item[] 11 = ULA
\item[] 12 = UART 0 Tx
\item[] 13 = UART 1 Tx (lowest priority)
\end{itemize}

\section{Interrupt Modes}
\paragraph{IM0}
When an interrupt is received by the CPU it disables interrupts and
executes the instruction placed on the bus by the interrupting device
and (no known use on the Next) It is enabled with the IM0 instruction
and enabling interrupts (EI).

\paragraph{IM1}
When an interrupt is received, the CPU disables interrupts and jumps
to an interrupt handler at \$0038 (normally in ROM). The ROM interrupt
handler updates the frame counter and scans the keyboard. This is the
default interrupt handling method for the ZX Spectrum and is probably
the method to use if you don’t need the ROMs for anything. It is
enabled using the IM1 instruction and enabling interrupts.

\paragraph{IM2}
The ZX Spectrum Next has both a legacy method for handling IM2 and an
updated one which makes better use of the capabilities of IM2 which
was added in Core 3.01.09.

The ZX Spectrum Next has 14 interrupt devices which can all be given
independent interrupt vectors when using IM2. These interrupts are
controlled by nextregs \$C0 -- \$CF.  The address of the vector for a
given interrupt is created by composing the I register (bits 15-0),
nextreg \$C0 bits 7-5 (bits 7-5) and the interrupt number of the
interrupt device (bits 4-1). This means that even if you use all 14
internal interrupt sources, your interrupt vector table is no more
than 28 bytes which can be at any 32 byte boundry. It also means that
far less processing has to be done on interrupts which are
received. External interrupts are a little different. If no vector is
supplied by a device the implied LSB will be \$FF.

While in IM2 mode, it is possible for interrupts to interrupt DMA
transfers. This capability is controlled by The DMA Interrupt enable
registers (nextregs \$CC -- \$CE). When DMA is interrupted, one
instruction of the main program will be processed, then the interrupt
will be taken. On return, DMA will continue.

\input{registers/tbblue/CC.tex}
\input{registers/tbblue/CD.tex}
\input{registers/tbblue/CE.tex}

In legacy mode, when the CPU receives an interrupt it disables
interrupts and jumps to an interrupt routine starting at the contents
of the jump table at I. The start of the interrupt routine is the
contents of I*\$100+bus and I*\$100+bus+1. Most devices that can
supply interrupts on the ZX Spectrum leave the data bus in a floating
state.  As a result the interpreted state of the data bus while
generally \$FF is not entirely predictable.  The solution to place
your interrupt routine at an address where the MSB and LSB are the
same (\$0101, \$0202, … \$FFFF) then place 257 copies of that value in
a block starting at I*\$100 (you can set the value of the I register).

Code:
\lstinputlisting{interrupts/interrupt.asm}

\input{interrupts/ctc.tex}
