M\chapter{memory}
Development
	
\textbf{Consider the memory location in the system. }

ZX Evolution has 4MB of memory.

The organization of this memory is similar to zx spectrum 128 - paging
is used.

Transcribed into pages of memory, we have 256 pages of 16 kb
each.

The above picture describes how the memory of the Spectrum is
organized. A similar system is used in TSconf, but somewhat expanded -
we are able to place pages not only from the address \#c000:
\begin{verbatim}\texttt{
Page0 (port #10af) - address #0000
Page1 (port #11af) - address #4000
Page2 (port #12af) - address #8000
Page3 (port #13af) - address #c000}
\end{verbatim}
Writing to the port switches the page in the window we need.

To use the paging system from address \#0000, you need to use the
MemConfig port (\#21af):

\begin{quotation}\textbf{Port MemConfig, \#21af:}
 
MemConfig LCK128 [1: 0] - - W0\_RAM! W0\_MAP W0\_WE ROM128

bits 7.6 - LCK128 - memory mode: 00-512k, 01 - 128k, 10 - Auto, 11 -
1024 k. The default is data from the BIOS.

bit 3 - W0\_RAM, 2 - W0\_MAP, 1 - W0\_WE, bit 0 - ROM128 (the same as
bit 4 in \#7FFD)

W0\_RAM 0 - ROM / 1 - RAM

! W0\_MAP 0 - mapped / 1 - Page0 

W0\_WE 0 - WP / 1 - WE (for ROM and RAM)

\end{quotation}

With its help, it is possible to switch the 0th memory window to read
/ write mode (the usual state is read only, there is the specter ROM),
and to enable mapping - the ability to place other pages of memory,
not just the standard page 0.

In window \#0000 you can enable RAM instead of ROM with optional
prohibition of recording in this window.

W0\_RAM = 0 - enabled ROM

W0\_RAM = 1 - enabled RAM 

Bit W0\_WE allows writing for RAM and programming (WE signal for ROM):

W0\_WE = 0 

W0\_RAM = 0 - ROM firmware is disabled

W0\_RAM = 1 - writing to RAM at \#0000-3FFF not allowed

W0\_WE = 1 

W0\_RAM = 0 - ROM firmware is allowed

W0\_RAM = 1 - writing to RAM at \#0000-3FFF is allowed

\textbf{Bit W0\_MAP} sets the page selection method for window \#0000.

 

W0\_MAP = 0 - the page number is selected depending on the current ZX
mode (Basic 48, Basic 128, TR-DOS).

Page layout: 0 - system, 1 - TR-DOS, 2 - Basic 128, 3 - Basic 48.

However, the register Page0 selects a bank of 4 pages to be used:

W0\_RAM = 0 - for window \#0000 is selected ROM

Page0 = 0bxxx000xx - pages 0-3 are used,

Page0 = 0bxxx001xx - pages 4-7 are used, 

... 

Page0 = 0bxxx111xx - pages 28-31 are used.

W0\_RAM = 1 - RAM is selected for the window \#0000 - this mode can be
used to debug the firmware without reprogramming the ROM or testing
non-standard pzushek.

Page0 = 0b000000xx - pages 0-3 are used, 

Page0 = 0b000001xx - pages 4-7 are used,

... 

Page0 = 0b111111xx - pages 252-255 are used.

W0\_MAP = 1 - the page number is taken directly from the register
Page0 - linear mode

W0\_RAM = 0 - pages

0-31 ROM W0\_RAM = 1 - pages 0-255 RAM

\textbf{Example:}

Turn on page \#80 from address \#8000
\begin{verbatim}\texttt{
		ld a,#80 ld bc,PAGE2 out (c),a}
\end{verbatim}
Enable the ability to work in page 0
\begin{verbatim}\texttt{
		ld bc,MemConfig ld a,%00001110 out (c),a}
\end{verbatim}
Voila, write, post with 0000.
