\chapter{Introduction}
The ZX Spectrum Next is an extension of the original ZX Spectrum
implemented in FPGA which implements many of the common additions to
the system includin the characteristics of all of the original ZX
Spectrum line, including the Timex/Sinclair 2068, along with a number
of characteristics to modernize the design.

This document is an attempt to consolidate the programming interface
for the ZX Spectrum Next into a single location. This document started
when much of the documentation on the ZX Spectrum Next site
(https://www.specnext.com/) was out of date and/or difficult to figure
out. The way to figure out how things actually worked was to either
dig through the forums and ask questions or find someones code that
implemented a particular bit of functionality and reverse engineer
it. The situation has greatly improved and this document may even be
redundant at this point.

Description from http://www.specnext.com/about/:

The Spectrum Next is fully implemented with FPGA technology, ensuring
it can be upgraded and enhanced while remaining truly compatible with
the original hardware by using special memory chips and clever
design. Here’s what under the hood of the machine:

\begin{itemize}
\item Processor: Z80 normal and turbo modes
\item Memory: 1024Kb RAM (expandable to 2048Kb on board)
\item Video: Hardware sprites, 256 colours mode and more.
\item Video Output: RGB, VGA, HDMI
\item Storage: SD Card slot, with DivMMC-compatible protocol
\item Audio: Turbo Sound Next (3x AY-3-8912 audio chips with stereo
  output)
\item Joystick: DB9 compatible with Cursor, Kempston and Interface 2
  protocols (selectable)
\item PS/2 port: Mouse with Kempston mode emulation and an external keyboard
\item Special: Multiface functionality for memory access, savegames,
  cheats etc.
\item Tape support: Mic and Ear ports for tape loading and saving
\item Expansion: Original external bus expansion port and accelerator
  expansion port
\item Accelerator board (optional): GPU / 1Ghz CPU / 512Mb RAM
\item Network (optional): Wi Fi module
\item Extras: Real Time Clock (optional), internal speaker (optional)
\end{itemize}

