\chapter{Basic Input/Output}

The basic I/O (human interface) system of the ZX Spectrum Next
supports keyboard, mouse and game controllers. These all extend the
functionality found on common ZX Spectrum peripherals. game
controllers and PS/2 keyboards can be customized to simulate responses
on the stock keyboard.

\section{Keyboard}

The ZX Spectrum Next can use a classic ZX Spectrum keyboard with a
$8\times 5$ matrix, an extended ZX Spectrum Next keyboard with a
$8\times 7$ matrix (the one used in a cased Next or the custom cases
for the N-Go), or a PS/2 keyboard. In all of these cases the system
translates the physical signals to look like a classic Spectrum
keyboard with access to the additional lines of the ZX Spectrum Next
keyboard as well.

The classic Spectrum part of the matrix is accessed using port \$xxFE
where there is a single unset bit in ``xx'' to select the row being
read. Next extensions are read using Nextregs \$B0 and \$B1 to read
columns 5 and 6.

\begin{center}
  \begin{tabular}{ | c || c | c | c | c | c | c | c || c | }
    \hline
    Bit & 0 & 1 & 2 & 3 & 4 & \$B0 & \$B1 & b \\
    \hline
    \$FEFE & Cap Sh & Z & X & C & V & $\Rightarrow$ & Extend & 0 \\
    \$FDFE & A & S & D & F & G & $\Leftarrow$ & Cap Lk & 1 \\
    \$FBFE & Q & W & E & R & T & $\Downarrow$ & Graph & 2 \\
    \$F7FE & 1 & 2 & 3 & 4 & 5 & $\Uparrow$ & True Vid & 3 \\
    \$FBFE & 0 & 9 & 8 & 7 & 6 & . & Inv Vid & 4  \\
    \$EFFE & P & O & I & U & Y & , & Break & 5  \\
    \$DFFE & Enter & L & K & J & H & \textquotedbl & Edit & 6 \\
    \$BFFE & Space & Sym Sh & M & N & B & ; & Del & 7 \\
    \hline
  \end{tabular}
\end{center}

\input{ports/7FFE.tex}
\input{ports/BFFE.tex}
\input{ports/DFFE.tex}
\input{ports/EFFE.tex}
\input{ports/F7FE.tex}
\input{ports/FBFE.tex}
\input{ports/FDFE.tex}
\input{ports/FEFE.tex}
\input{registers/tbblue/B0.tex}
\input{registers/tbblue/B1.tex}

\section{Game Controllers}

The ZX Spectrum used a number of joystick standards and the ZX
Spectrum Next can make Atari controllers or Mega Drive controllers
look like many of these standards. Atari joystick and driving (not the
similar looking paddle) controllers are supported by the
interface. The interface supports Mega Drive controllers with up to 11
buttons (start, a, b, c, x, y, z, up, down, left, right). It is
possible to map unused buttons to keys and simulate unavailable
buttons with keys.

Nextreg \$05 selects the interface mode for each of the two
joysticks. If in Kempston or Megadrive mode the joystick 1 can be read
using port \$1F and joystick 2 using port \$37. When using these ports
in Megadrive mode, some buttons will be paired A/X, B/Y, C/Z with B/Y
also corresponding to the single fire button on Atari joysticks. In
order to disambiguate presses of A/X, B/Y, and C/Z it is necessary to
also read nextreg \$B2.

\input{ports/1F.tex}
\input{ports/37.tex}
\input{registers/tbblue/05.tex}
\input{registers/tbblue/B2.tex}

\section{Mouse}

A mouse attached to the PS/2 port looks like a kempston mouse to the
system. Port \$FADF is used to read the state of the mouse wheel and
buttons while ports \$FBDF and \$FBDF are used to read the X and Y
positions.

\input{ports/FADF.tex}
\input{ports/FBDF.tex}
\input{ports/FFDF.tex}

\section{Keymapping}

Both a PS/2 keyboard and the contoller buttons can be given custom
keyboard mappings. These mappings are accomplished using nextregs \$28
(keymap address MSB), \$29 (keymap address LSB), \$2A (keymap data
MSB) and \$2B (keymap data LSB).

\input{registers/tbblue/28.tex}
\input{registers/tbblue/29.tex}
\input{registers/tbblue/2A.tex}
\input{registers/tbblue/2B.tex}

\subsection{Keyboard}

To remap PS/2 keyboard keys, the starting address for the keymap data
is written to nextregs \$28 and \$29 keyboard data address followed by
writing the data to nextregs \$2A and \$2B with the write to \$2B
autoincrementing the address for the data.

\subsection{Joystick}

To map joystick buttons an \$80 is written to nextreg followed by a
\$00 (left) or \$10 (right) to nextreg \$29 to select which joystick
is being mapped to the keyboard 11 writes to nextgreg \$2B to indicate
the key(s) for each button. Each byte to addresses the row and column
of the ZX Next keyboard with bits 5-3 being the row and bits 2-0 being
the column (column 7, 111 is no action).
